---
title: "simple-EDA"
author: "Helen Evangelina"
date: "28/09/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}
training <- read_csv("data/training.csv")
training <- training %>%
  filter(!CAUSE %in% c("BURNING BUILDING",
                       "WASTE DISPOSAL, INDUSTRIAL, SAWMILL, TIP",
                       "WASTE DISPOSAL, DOMESTIC",
                       "BURNING VEHICLE, MACHINE",
                       "BURNING BUILDING")) %>%
  filter(new_cause != "other") %>%
  filter(new_cause != "relight")
training <- dplyr::select(training, -c(EVENTID:FIRE_NUM), -id, -CAUSE, -FOR_CODE)
training <- training %>% 
  mutate(year = factor(lubridate::year(FIRE_START)),
         month = factor(lubridate::month(FIRE_START)),
                        # levels = c(10,11,12,1,2,3),
         day = factor(lubridate::day(FIRE_START), 
                      levels = c(1:31)),
         wod = factor(lubridate::wday(FIRE_START), 
                      levels = c(1:7)))
# training <- training %>% 
#   filter(month %in% c(10,11,12,1,2,3))
training <- na.omit(training)
training <- training %>% 
  mutate(new_cause = ifelse(test = new_cause == "accidental_human",
                            yes = "accident",
                            no = new_cause)) %>%
  mutate(new_cause = ifelse(test = new_cause == "burning_off_human", 
                            yes = "burning_off", 
                            no = new_cause)) %>%
  mutate(new_cause = factor(new_cause)) %>%
  mutate(FOR_TYPE = factor(FOR_TYPE))
training <- na.omit(training)
training <- training %>% 
  mutate(log_dist_cfa = log(dist_cfa),
         log_dist_camp = log(dist_camp),
         log_dist_road = log(dist_road),
         COVER = factor(COVER),
         HEIGHT = factor(HEIGHT))
training <- training %>% 
  rename(cause = new_cause)
training <- mutate(training,
                   cause = forcats::fct_relevel(cause,
                                                "lightning",
                                                "accident",
                                                "arson",
                                                "burning_off"))
training <- na.omit(training)
training <- training %>% 
  dplyr::select(-dist_road, -dist_cfa, -dist_camp, -FIRE_START)
```

```{r}
# Read in predict data
predict_x <- readr::read_csv("data/predict_x.csv")
predict_x <- predict_x %>% 
  mutate(log_dist_cfa = log(dist_cfa),
         log_dist_camp = log(dist_camp),
         log_dist_road = log(dist_road))
```

# simple EDA
## for the training data (less reliable)
## this is to see if vegetation category has an effect on the fire risk
```{r}
FOR_cat_COUNT <- training %>%
  group_by(FOR_CAT, year)%>%
  summarise(n = n())

FOR_cat_COUNT %>%
  ggplot(aes(x = reorder(FOR_CAT, -n),
             y = n)) +
  geom_boxplot() +
  xlab("FOR_CAT") +
  ylab("Count") +
  theme_bw() +
  ggtitle("Training data")
```



```{r}
# Read in predict data
predict_x <- readr::read_csv("data/predict_x.csv")
predict_x <- predict_x %>% 
  mutate(log_dist_cfa = log(dist_cfa),
         log_dist_camp = log(dist_camp),
         log_dist_road = log(dist_road))
```

## on the "predict" dataset - from hotspot data (more reliable)
```{r}
FOR_cat_COUNT2 <- predict_x %>%
  group_by(FOR_CAT, year)%>% 
  na.omit() %>%
  summarise(n = n())

FOR_cat_COUNT2 %>%
  ggplot(aes(x = reorder(FOR_CAT, -n),
             y = n)) +
  geom_boxplot() +
  xlab("FOR_CAT") +
  ylab("Count") +
  theme_bw() +
  ggtitle("Predict data")
```

Conclusion: 
The "training" and "predict" data show quite different results. Native forest is the category in where most fires happened in both training and predict dataset, followed by non-forest, commercial plantation and other forest. (THE ORDER ARE THE SAME). Just, in 2019-2020 there is a lot of fires in non-forest. 

--> Vegetation type has an impact

## look at other variables
## COVER
```{r}
#training data
FOREST_COUNT <- training %>%
  group_by(FOREST, year)%>% 
  na.omit() %>%
  summarise(n = n())

FOREST_COUNT %>%
  ggplot(aes(x = reorder(FOREST, -n),
             y = n)) +
  geom_boxplot() +
  xlab("FOREST") +
  ylab("Count") +
  theme_bw() +
  ggtitle("Training data")

```

```{r}
#predict data
FOREST_COUNT2 <- predict_x %>%
  group_by(FOREST, year)%>% 
  na.omit() %>%
  summarise(n = n())

FOREST_COUNT2 %>%
  ggplot(aes(x = reorder(FOREST, -n),
             y = n)) +
  geom_boxplot() +
  xlab("FOREST") +
  ylab("Count") +
  theme_bw() +
  ggtitle("Predict data")

```

Conc: forest has an effect --> forest areas are more likely to have fires. Diff: the gap in predict dataset is larger. 

## COVER
```{r}
#training data
COVER_COUNT <- training %>%
  group_by(COVER, year)%>% 
  na.omit() %>%
  summarise(n = n())

COVER_COUNT %>%
  ggplot(aes(x = reorder(COVER, -n),
             y = n)) +
  geom_boxplot() +
  xlab("COVER") +
  ylab("Count") +
  theme_bw() +
  ggtitle("Training data")

```

```{r}
# PREDICT data
COVER_COUNT2 <- predict_x %>%
  group_by(COVER, year)%>% 
  na.omit() %>%
  summarise(n = n())

COVER_COUNT2 %>%
  ggplot(aes(x = reorder(COVER, -n),
             y = n)) +
  geom_boxplot() +
  xlab("COVER") +
  ylab("Count") +
  theme_bw() +
  ggtitle("Predict data")

```

Conc: same order as well but predict has higher gap. 
also, one outlier in the training data for 2.

## HEIGHT
```{r}
#training data
HEIGHT_COUNT <- training %>%
  group_by(HEIGHT, year)%>% 
  na.omit() %>%
  summarise(n = n())

HEIGHT_COUNT %>%
  ggplot(aes(x = reorder(HEIGHT, -n),
             y = n)) +
  geom_boxplot() +
  xlab("HEIGHT") +
  ylab("Count") +
  theme_bw() +
  ggtitle("Training data")

```

```{r}
# predict data
HEIGHT_COUNT2 <- predict_x %>%
  group_by(HEIGHT, year)%>% 
  na.omit() %>%
  summarise(n = n())

HEIGHT_COUNT2 %>%
  ggplot(aes(x = reorder(HEIGHT, -n),
             y = n)) +
  geom_boxplot() +
  xlab("HEIGHT") +
  ylab("Count") +
  theme_bw() +
  ggtitle("predict data")

```
same conc. 




